---
title: "ggfoundry"
description: | 
  Add a layer of custom fillable shapes to a ggplot.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggfoundry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

## Motivation

ggfoundry was inspired a little by Stack Overflow posts seeking specific shapes. But, in truth, mostly by a personal interest in getting acquainted with grid graphics (the underpinnings of ggplot2).

## Shape landscape

Yes, there is already a seemingly near-infinite number of shapes out there:

-   Those familiar to ggplot users (some fillable) as described in the [ggplot2 documentation](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html#point);
-   Colourable [unicodes](https://www.compart.com/en/unicode/category/So) and [icons](https://fontawesome.com/icons/) like fontawesome;
-   [ggimage](https://github.com/GuangchuangYu/ggimage) enables the use of whole pictures;
-   And then there is the DIY (Do-It-Yourself) approach: Conjuring up grobs (grid graphical objects); perhaps with a sprinkle of trigonometry.

But sometimes you just can't find what you want. Nor manipulate it in the way you would like.

ggfoundry offers arbitrary hand-crafted colourable and fillable shapes for ggplot2 and is reviewed side-by-side with other options in [contrast with alternatives](https://cgoo4.github.io/ggfoundry/articles/contrast).

## Foundry process

These artisanal symbols begin life as vector images with two layers: an outline and a fill. Some are hand-drawn; others, such as the "bowl" set, are constructed programmatically. Each SVG pair is converted to Cairo graphics format, forged at extreme temperatures into objects of class "Picture", and finally delicately cast as a `gTree` representation of the original shape. But not quite back to where we started, because they are now editable.

When cooled and finely burnished, the `gTree` and all its grob children may then be manipulated by `geom_casting()` to render the desired ggplot with those special high-end adornments.

## Available shapes

ggfoundry may well be the destination of "last resort"!

After travelling the mountains, seas and forests of the world in search of that elusive shape (or small set), a hand-made grob may be the fillable "Holy Grail" sought via a [Github issue](https://github.com/cgoo4/ggfoundry/issues).

These sets are included with the latest version of the package. You can "mix and match" shapes from different sets; the "set" is for grouping shapes in the documentation and for use in `shapes_cast()` to filter for the desired shapes.

```{r sets, message=FALSE, warning=FALSE, fig.height=7}
library(ggfoundry)
library(dplyr)
library(forcats)
library(stringr)

df <- shapes_cast() |> 
  filter(!str_ends(shape, "3|4|5|6")) |> 
  mutate(x = row_number(), shape = fct_inorder(shape), .by = set)

df |> 
  ggplot(aes(x, set)) +
  geom_text(aes(label = shape), nudge_y = -0.5, colour = "grey70", size = 3) +
  geom_casting(aes(shape = shape), size = 0.19, fill = "skyblue") +
  scale_shape_manual(values = as.character(df$shape)) +
  scale_x_continuous(expand = expansion(add = 0.5)) +
  scale_y_discrete(expand = expansion(add = 0.7)) +
  labs(x = NULL, y = NULL, caption = "sunflowers 1-8 available") +
  theme_minimal() +
  theme(
    text = element_text(colour = "grey70"),
    axis.text.y = element_text(angle = 90, hjust = 0.5),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
    )
```

## Simple example

Using some made-up data simulating a "random walk", `geom_casting()` adds a layer of custom shapes to the plot.

When the shape is mapped to a variable, then `scale_shape_manual()` is required to explicitly name the desired shapes as a character vector. This is because standard shapes (as used for example in `geom_point()`) are associated with a number, e.g. a circle is 19, whereas `geom_casting()` shapes are associated only with character strings.

Each shape's appearance is resolved observation by observation, so aesthetics such as `colour`, `fill`, `alpha`, `size` and `angle` may vary continuously across the data, whether grouping is implicit or explicitly shared with `group = 1`.

Using `scale_colour_manual()` and `scale_fill_manual()`, we can also select a custom palette for the shape colours and fills.

```{r}
# Toy Data
set.seed(123)

random_walk <- \(x, y, z) cumsum(rnorm(x, mean = y, sd = sqrt(z)))

df <- data.frame(
  x = rep(1:10, 2),
  y = c(
    random_walk(10, 1, 1),
    random_walk(10, 3, 1.3)
  ),
  group = factor(c(rep(1, 10), rep(2, 10)))
)

# Plot with geom_casting()
df |>
  ggplot(aes(x, y, shape = group, colour = group, fill = group)) +
  geom_line(show.legend = FALSE) +
  geom_casting() +
  scale_colour_manual(values = c("darkred", "darkgreen")) +
  scale_fill_manual(values = c("pink", "lightgreen")) +
  scale_shape_manual(values = c("cross1", "cross2")) +
  labs(title = "ggfoundry") +
  theme_bw() +
  theme(plot.subtitle = element_text(size = 10))
```

## Porridge bowls

The "bowl" set brings three containers to the table: `bowl0`, `bowl1` and `bowl2`, distinguished only by zero, one or two wisps of steam rising from the porridge. The bowl body and the steam share the `fill`, while the bowl and porridge outline take the `colour`. All three variants share the same canvas and bowl position, so they may be swapped freely with `scale_shape_manual()`.

A continuous `fill` maps directly to any numeric variable, with no grouping workaround required. And a `fill` of `"transparent"` hides the steam while the bowl outline remains visible.

```{r bowls, message=FALSE, warning=FALSE, fig.width=11, fig.height=4}

bowls <- data.frame(
  x = 1:7,
  temperature = c(5, 18, 32, 47, 63, 81, 96)
)

bowls |>
  ggplot(aes(x, 1, fill = temperature)) +
  geom_casting(shape = "bowl2", colour = "#272626", size = 0.72) +
  geom_text(aes(y = 0.35, label = temperature), size = 4) +
  scale_fill_viridis_c(limits = c(0, 100), begin = 0.04, end = 0.94) +
  scale_x_continuous(limits = c(0.35, 7.65), expand = expansion(mult = 0)) +
  scale_y_continuous(limits = c(0, 1.85), expand = expansion(mult = 0)) +
  labs(
    title = "Seven values; one shape; no grouping workaround",
    fill = "Value"
  ) +
  theme_void() +
  theme(legend.position = "bottom")
```

See the [showcase](https://cgoo4.github.io/ggfoundry/articles/example_uses) article to explore other use cases and [contrast with alternatives](https://cgoo4.github.io/ggfoundry/articles/contrast) to review against other options.

## Acknowledgements

Without the pivotal grConvert [@grConvert] and grImport2 [@grImport2] packages, the foundry process would not have been viable. And only thanks to the work behind ggplot2 [@ggplot2], may the shapes cast be so beautifully visualised.
